Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2022-09-27 analysis/paper_draft3.Rmd
Update paper_draft3.Rmd
revision edits (equations, b_hist)

---
title: "Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data"
date: "`r Sys.Date()`"
blinded: 0
authors:
- name: "[Sachin S. Pandya](https://orcid.org/0000-0001-7387-1307)"
  #thanks: The authors gratefully acknowledge ...
  affiliation: University of Connecticut
  email: sachin.pandya@uconn.edu
- name: "Xiaomeng Li"
  affiliation: University of Connecticut
- name: "[Eric Barón](https://orcid.org/0000-0003-1934-1932)"
  affiliation: University of Connecticut
- name: "[Timothy E. Moore](https://orcid.org/0000-0002-9576-0517)"
  affiliation: University of Connecticut
output:
  bookdown::pdf_book: 
      base_format: rticles::asa_article
      citation_package: "default"
      keep_tex: TRUE
header-includes: 
  - \usepackage{amsmath}
#fontsize: 12pt
bibliography: references.bib
biblio-style: agsm.bst
nocite: |
  @WashR37, @calccp2021
keywords: "Batson challenge, peremptory strikes, power prior, Bayesian"
abstract: |
  US law bars using peremptory strikes during jury selection because of prospective juror race, ethnicity, gender, or membership in certain other cognizable classes. Here, we extend a Bayesian approach for detecting such illegal bias. We show how to incorporate historical data on an attorney's use of peremptory strikes in past cases to estimate how likely that attorney has struck prospective jurors in a current trial because of their cognizable-class membership. In so doing, we use the power prior to adjust the weight of such historical information in the analysis. We show how well our model performs on simulated data. Finally, we extend this approach with a prototype software application to detect bias in peremptory-challenge in real time during jury-selection. We illustrate this application's use with historical strike data from a convenience sample of cases from one court.

@@

Since _Batson v. Kentucky_ [-@batson1986], a party violates the Equal Protection Clause of the US Constitution by using peremptory challenges if motivated by the prospective jurors' race, ethnicity, gender, or certain other cognizable classes. The party bringing a _Batson_ challenge bears the burden of proving such illegal bias is more likely than not to be true [@LaFave2022, $\S$ 22.3(d)]. For similar challenges under State law, a few States require only that an "objective observer" or an "objectively reasonable person" would find that race, ethnicity or another cognizable-class was a "factor" in that party's use of strikes [Wash. General Rule. 37(e); Calif. Code of Civil Procedure $\S$ 231.7(d)(1)]. Peremptory-challenge procedure, and thus the task of proving such illegal bias, varies not only by State, but also by court, including the number of peremptory challenges assigned to each side and the order in which each party uses those strikes [@nscs-jury2018; @Williams2017].

In any case, evidence of such illegal bias may include data on a party's use of peremptory challenges in past cases [@flowers2019, 2243; Wash. General Rule 37(g)(v); Calif. Code Civil Procedure $\S$ 231.7(d)(3)(G)]. Prior studies have collected historical strike data in past cases and reported the observed difference in strike rates by race or gender. Typically, they test for the probability of observing a non-zero difference in strike rates by the race or gender of the struck prospective jurors, given repeated sampling from a hypothetical population of peremptory strikes with zero such difference [e.g., @Eisenberg2017; @Grosso2012; for discussion, see @Gastwirth2014, 289-297; @Gastwirth2013]. Prior studies have also modeled how much a prospective juror's race affected the odds of being struck. For different modeling approaches using the same historical strike data from Mississippi, see @Craft2018a; @DeCamp2021; and @Dunn2021.

@@

Accordingly, for any given case $i$ in which jury selection occurs, let $j$ denote a peremptory strike used, and let $\delta_{ij}$ denote whether or not a party used that strike on a person who belongs to a "cognizable class". If "race" is the bias type of interest, the cognizable class is racial minority jurors ($\delta_{ij} =1$, 0 for White jurors).  If "gender" is the bias of interest, the cognizable class is female jurors ($\delta_{ij} = 1$, 0 for male jurors). In turn, let $c_{ij}$ denote the number of cognizable class members subject to strike; and $m_{ij}$ denote the number of cognizable class non-members subject to strike, such that $c_{ij}+m_{ij}$ is the total number of jurors potentially subject to strike. If there is no bias, the probability is $\frac{c_{ij}}{c_{ij}+m_{ij}}$ for striking a cognizable class member, and $1-\frac{c_{ij}}{c_{ij}+m_{ij}}$ for striking someone who does not belong to that class.

@@

\begin{equation}  
\pi(b|D_0,\alpha_0) \propto L(b|D_0)^{\alpha_0}\pi_0(b)
\label{eq:powerprior}
\end{equation} where $0 \le \alpha_0 \le 1$ is the parameter controlling the weight of the historical information; $D_0$ is the observed historical data; $L(b|D_0)$ is the likelihood function of $b$ given the historical data; and $\pi_0(b)$ is the initial prior before the historical data is observed.

@@

\begin{equation}
  L(b|\delta) \propto L(b|\delta)(L(b|\delta_0))^{\alpha_0}\exp(-\frac{(b-\mu)^2}{2\sigma^2})
  \label{eq:model4}
\end{equation} where

@@

We generated historical strike data of three sizes (same, double, and triple the size of
current data, i.e., data on one previous trial, two previous trials) using seven different values of the bias parameter for generating that historical data $b_{hist}=\{-3,-2,-1,0,1,2,3\}$.

>>>>

---
title: "Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data"
date: "`r Sys.Date()`"
blinded: 0
authors:
- name: "[Sachin S. Pandya](https://orcid.org/0000-0001-7387-1307)"
  #thanks: The authors gratefully acknowledge ...
  affiliation: University of Connecticut
  email: sachin.pandya@uconn.edu
- name: "Xiaomeng Li"
  affiliation: University of Connecticut
- name: "[Eric Barón](https://orcid.org/0000-0003-1934-1932)"
  affiliation: University of Connecticut
- name: "[Timothy E. Moore](https://orcid.org/0000-0002-9576-0517)"
  affiliation: University of Connecticut
output:
  bookdown::pdf_book: 
      base_format: rticles::asa_article
      citation_package: "default"
      keep_tex: TRUE
header-includes: 
  - \usepackage{amsmath}
#fontsize: 12pt
bibliography: references.bib
biblio-style: agsm.bst
nocite: |
  @WashR37, @calccp2021
keywords: "Batson challenge, peremptory strikes, power prior, Bayesian"
abstract: |
  US law bars using peremptory strikes during jury selection because of prospective juror race, ethnicity, sex, or membership in certain other cognizable classes. Here, we extend a Bayesian approach for detecting such illegal bias. We show how to incorporate historical data on an attorney's use of peremptory strikes in past cases to estimate how likely that attorney has struck prospective jurors in a current trial because of their cognizable-class membership. In so doing, we use the power prior to adjust the weight of such historical information in the analysis. We show how well our model performs on simulated data. Finally, we extend this approach with a prototype software application to detect bias in peremptory-challenge in real time during jury-selection. We illustrate this application's use with historical strike data from a convenience sample of cases from one court.

@@

Since _Batson v. Kentucky_ [-@batson1986], a party violates the Equal Protection Clause of the US Constitution by using peremptory challenges if motivated by the prospective jurors' race, ethnicity, sex, or certain other cognizable classes. The party bringing a _Batson_ challenge bears the burden of proving such illegal bias is more likely than not to be true [@LaFave2022, $\S$ 22.3(d)]. For similar challenges under State law, a few States require only that an "objective observer" or an "objectively reasonable person" would find that race, ethnicity or another cognizable-class was a "factor" in that party's use of strikes [Wash. General Rule. 37(e); Calif. Code of Civil Procedure $\S$ 231.7(d)(1)]. Peremptory-challenge procedure, and thus the task of proving such illegal bias, varies not only by State, but also by court, including the number of peremptory challenges assigned to each side and the order in which each party uses those strikes [@nscs-jury2018; @Williams2017].

In any case, evidence of such illegal bias may include data on a party's use of peremptory challenges in past cases [@flowers2019, 2243; Wash. General Rule 37(g)(v); Calif. Code Civil Procedure $\S$ 231.7(d)(3)(G)]. Prior studies have collected historical strike data in past cases and reported the observed difference in strike rates by race or sex. Typically, they test for the probability of observing a non-zero difference in strike rates by the race or sex of the struck prospective jurors, given repeated sampling from a hypothetical population of peremptory strikes with zero such difference [e.g., @Eisenberg2017; @Grosso2012; for discussion, see @Gastwirth2014, 289-297; @Gastwirth2013]. Prior studies have also modeled how much a prospective juror's race affected the odds of being struck. For different modeling approaches using the same historical strike data from Mississippi, see @Craft2018a; @DeCamp2021; and @Dunn2021.

@@

Accordingly, for any given case $i$ in which jury selection occurs, let $j$ denote a peremptory strike used, and let $\delta_{ij}$ denote whether or not a party used that strike on a person who belongs to a "cognizable class". If "race" is the bias type of interest, the cognizable class is racial minority jurors ($\delta_{ij} =1$, 0 for White jurors).  If "sex" is the bias of interest, the cognizable class is female jurors ($\delta_{ij} = 1$, 0 for male jurors). In turn, let $c_{ij}$ denote the number of cognizable class members subject to strike; and $m_{ij}$ denote the number of cognizable class non-members subject to strike, such that $c_{ij}+m_{ij}$ is the total number of jurors potentially subject to strike. If there is no bias, the probability is $\frac{c_{ij}}{c_{ij}+m_{ij}}$ for striking a cognizable class member, and $1-\frac{c_{ij}}{c_{ij}+m_{ij}}$ for striking someone who does not belong to that class.

@@

\begin{equation}  
\pi(b|\delta_0,\alpha_0) \propto L(b|\delta_0)^{\alpha_0}\pi_0(b)
\label{eq:powerprior}
\end{equation} where $0 \le \alpha_0 \le 1$ is the parameter controlling the weight of the historical information; $\delta_0$ is the observed historical data; $L(b|\delta_0)$ is the likelihood function of $b$ given the historical data; and $\pi_0(b)$ is the initial prior before the historical data is observed.

@@

\begin{equation}
  P(b|\delta,\delta_0) \propto L(b|\delta)(L(b|\delta_0))^{\alpha_0}\exp(-\frac{(b-\mu)^2}{2\sigma^2})
  \label{eq:model4}
\end{equation} where

@@

We generated historical strike data of three sizes (same, double, and triple the size of
current data, i.e., data on one previous trial, two previous trials) using seven different values of the bias parameter for generating that historical data $b_{hist}=\{0,0.5,1,1.5,2, 2.5,3\}$.
